---
title: "Democratic Backsliding and Foreign Aid"
author: "Sania Bala"
date: today
format: html
---

## Setup

Loading required packages. Any missing packages are installed automatically.

```{r}
#| message: false
#| warning: false

required_packages <- c("tidyverse", "WDI", "countrycode", "MASS", "here")
missing_packages <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if (length(missing_packages) > 0) install.packages(missing_packages, repos = "https://cloud.r-project.org")

library(tidyverse)
library(WDI)
library(countrycode)
library(here)
# MASS is loaded later, just before the models, to avoid it masking dplyr::select
```

## Data

### Load raw data

We use two datasets:

-   **V-Dem** (Varieties of Democracy): country-year measures of democracy, governance, and civil liberties
-   **OECD DAC2A**: total ODA (Official Development Assistance) disbursements by recipient country and year, 1960--2024

```{r}
#| message: false

vdem <- read_csv(here("VDEM.csv"))
oecd_raw <- read_csv(here("sania/full_oecd_D.csv"))
```

### Clean and aggregate OECD data

Country names differ between the two datasets, so we standardize OECD recipient names to match V-Dem before merging. We then aggregate from donor-level rows to total ODA received per recipient country per year.

```{r}
oecd_clean <- oecd_raw %>%
  mutate(Recipient = recode(Recipient,
    "Myanmar"                                           = "Burma/Myanmar",
    "Cabo Verde"                                        = "Cape Verde",
    "China (People's Republic of)"                      = "China",
    "Côte d'Ivoire"                                     = "Ivory Coast",
    "Lao People's Democratic Republic"                  = "Laos",
    "Democratic People's Republic of Korea"             = "North Korea",
    "Palestinian Authority or West Bank and Gaza Strip" = "Palestine/West Bank",
    "Congo"                                             = "Republic of the Congo",
    "Syrian Arab Republic"                              = "Syria",
    "Gambia"                                            = "The Gambia",
    "Viet Nam"                                          = "Vietnam"
  ))

oecd <- oecd_clean %>%
  group_by(Recipient, TIME_PERIOD) %>%
  summarise(total_oda_usd_millions = sum(OBS_VALUE, na.rm = TRUE), .groups = "drop") %>%
  rename(country_name = Recipient, year = TIME_PERIOD)

write_csv(oecd, here("sania/oecd_clean.csv"))
```

### Merge V-Dem and OECD

We merge on `country_name` and `year`, keeping all V-Dem observations (left join). We also compute `delta_polyarchy` --- the year-over-year change in the electoral democracy index --- **before** subsetting years, so that the first observation in the analysis window can draw its lag from the prior year.

-   Negative `delta_polyarchy` = democratic backsliding
-   Positive `delta_polyarchy` = democratization

```{r}
merged <- vdem %>%
  left_join(oecd, by = c("country_name", "year")) %>%
  arrange(country_name, year) %>%
  group_by(country_name) %>%
  mutate(delta_polyarchy = v2x_polyarchy - lag(v2x_polyarchy)) %>%
  ungroup()
```

### Subset to analysis variables (1960--2024)

We keep only the variables needed for analysis and filter to the period covered by the OECD data.

| Variable          | Description                                        |
|-------------------|----------------------------------------------------|
| `v2x_polyarchy`   | Electoral democracy level (control)                |
| `delta_polyarchy` | Year-over-year change in democracy (key predictor) |
| `v2x_corr`        | Political corruption index                         |
| `v2x_rule`        | Rule of law index                                  |
| `v2xcl_rol`       | Civil liberties / equality before the law          |

```{r}
analysis <- merged %>%
  filter(year >= 1960, year <= 2024) %>%
  dplyr::select(
    country_name, year, total_oda_usd_millions,
    v2x_polyarchy, delta_polyarchy,
    v2x_corr, v2x_rule, v2xcl_rol
  )
```

### Add GDP per capita and population

GDP per capita and population are not in our V-Dem extract, so we pull them from the World Bank via the `WDI` package.

```{r}
#| message: false
#| warning: false

wdi <- WDI(
  indicator = c(gdp_per_capita = "NY.GDP.PCAP.KD",
                population     = "SP.POP.TOTL"),
  start = 1960, end = 2024,
  extra = FALSE
) %>%
  mutate(country_name = countrycode(iso2c, origin = "iso2c",
                                    destination = "country.name")) %>%
  filter(!is.na(country_name)) %>%
  dplyr::select(country_name, year, gdp_per_capita, population)

analysis <- analysis %>%
  left_join(wdi, by = c("country_name", "year"))
```

### Prepare outcome variable

ODA is reported in USD millions (continuous). We convert it to an integer count by rounding to the nearest million and flooring any negative values (loan repayments) to zero, as required by count models.

```{r}
analysis <- analysis %>%
  mutate(oda_count = pmax(0, round(total_oda_usd_millions))) %>%
  dplyr::select(-total_oda_usd_millions) %>%
  filter(!is.na(oda_count), !is.na(v2x_polyarchy))


write_csv(analysis, here("sania/analysis_merged.csv"))
```

## Models

### Model strategy

We use negative binomial regression to test whether democratic backsliding predicts foreign aid received. ODA is heavily overdispersed --- its variance far exceeds its mean --- which rules out Poisson. Population is included as a control (`log(population)`).

```{r}
#| message: false
library(MASS)
```

### Overdispersion check

Under Poisson, the variance should equal the mean. If the variance-to-mean ratio is much greater than 1, the data is overdispersed and negative binomial is more appropriate.

```{r}
tibble(
  mean     = mean(analysis$oda_count, na.rm = TRUE),
  variance = var(analysis$oda_count,  na.rm = TRUE),
  ratio    = var(analysis$oda_count,  na.rm = TRUE) / mean(analysis$oda_count, na.rm = TRUE)
)
```

The variance-to-mean ratio is far greater than 1, confirming severe overdispersion. We use negative binomial regression, which adds a dispersion parameter to account for this.

### Backsliding model: Democratic backsliding and aid (Negative Binomial)

The key predictor is `delta_polyarchy` (year-over-year change in `v2x_polyarchy`). We keep `v2x_polyarchy` as a level control so that `delta_polyarchy` captures the effect of change net of overall democratic standing. The expected sign on `delta_polyarchy` is positive --- backsliding (negative delta) should predict less ODA.

```{r}
nb_backsliding <- glm.nb(
  oda_count ~ delta_polyarchy + v2x_corr + v2x_rule + v2xcl_rol +
              log(gdp_per_capita) + log(population),
  data = analysis
)

summary(nb_backsliding)

```

**Results:** n = 6,466 country-years (1,084 dropped due to missingness). Theta = 0.992, confirming overdispersion. Residual deviance = 7,525 on 6,458 df. AIC = 91,429.

| Predictor | Estimate | p-value | Interpretation |
|------------------|------------------|------------------|------------------|
| `delta_polyarchy` | −0.310 | .264 | **Not significant.** Direction reversed from prior specification --- backsliding weakly associated with less aid, but indistinguishable from zero |
| `v2x_polyarchy` | −0.435 | \<.001 | More democratic countries receive less ODA; consistent with need-based allocation |
| `v2x_corr` | 1.142 | \<.001 | Higher corruption associated with more ODA |
| `v2x_rule` | 1.616 | \<.001 | Better rule of law associated with more ODA |
| `v2xcl_rol` | 0.382 | \<.001 | Stronger civil liberties associated with more ODA |
| `log(gdp_per_capita)` | −0.245 | \<.001 | Richer countries receive less ODA |
| `log(population)` | 0.545 | \<.001 | Larger countries receive more ODA |

**Interpretation:** The backsliding hypothesis is not supported in this specification. Aid allocations respond to a country's overall democratic standing, but not to year-to-year democratic change. Notably, with the extended time series the direction of `delta_polyarchy` flips --- backsliding is now weakly associated with *less* aid (negative coefficient), which is the expected direction, but the effect remains statistically indistinguishable from zero. This likely reflects donor inertia --- aid budgets are planned in advance and respond slowly to political shifts.

## Robustness Checks

### Check 1: Alternative democracy measures

`v2x_polyarchy` measures electoral democracy specifically. If the null result is real, it should hold across different conceptualizations of democracy. We test two alternatives:

-   `v2x_libdem` --- liberal democracy (adds judicial constraints and civil liberties)
-   `v2x_partipdem` --- participatory democracy (emphasizes civil society and direct participation)

```{r}
analysis_robust <- merged %>%
  arrange(country_name, year) %>%
  group_by(country_name) %>%
  mutate(
    delta_libdem    = v2x_libdem - lag(v2x_libdem),
    delta_partipdem = v2x_partipdem - lag(v2x_partipdem)
  ) %>%
  ungroup() %>%
  filter(year >= 1960, year <= 2024) %>%
  dplyr::select(country_name, year, v2x_libdem, v2x_partipdem,
                delta_libdem, delta_partipdem) %>%
  left_join(analysis, by = c("country_name", "year"))

nb_libdem <- glm.nb(
  oda_count ~ delta_libdem + v2x_corr + v2x_rule + v2xcl_rol +
              log(gdp_per_capita) + log(population),
  data = analysis_robust
)

nb_partipdem <- glm.nb(
  oda_count ~ delta_partipdem + v2x_corr + v2x_rule + v2xcl_rol +
              log(gdp_per_capita) + log(population),
  data = analysis_robust
)

summary(nb_libdem)
summary(nb_partipdem)
```

**Results:** The null result holds across both alternatives.

| Model                                       | Delta estimate | p-value |
|---------------------------------------------|----------------|---------|
| Electoral democracy (`delta_polyarchy`)     | 0.466          | .537    |
| Liberal democracy (`delta_libdem`)          | 0.185          | .840    |
| Participatory democracy (`delta_partipdem`) | 1.326          | .266    |

The null result on backsliding is robust across all three conceptualizations of democracy. This consistently points to donor inertia --- aid allocations respond to where a country sits on the democracy scale, not to how fast it is moving.

### Check 2: Extended time window

The analysis now uses the full OECD time series (1960--2024), providing a much longer panel with more backsliding episodes and greater statistical power compared to the original 2010--2020 window. Year coverage in practice is limited by the availability of V-Dem scores and World Bank GDP/population data, but the full window is retained and listwise deletion handles missing observations.

## Replication: Commitment Data

ODA commitments capture when donors pledge aid, while disbursements capture when it is actually transferred. Commitments may respond more quickly to political changes since they are forward-looking decisions, whereas disbursements reflect prior commitments and are stickier. We replicate the main analysis using OECD DAC3A commitment data (1966--2024) to test whether the null result on backsliding holds.

### Load and clean commitment data

```{r}
oecd_commit_raw <- read_csv(here("sania/oecd_commitments.csv"))

oecd_commit <- oecd_commit_raw %>%
  mutate(Recipient = recode(Recipient,
    "Myanmar"                                           = "Burma/Myanmar",
    "Cabo Verde"                                        = "Cape Verde",
    "China (People's Republic of)"                      = "China",
    "Côte d'Ivoire"                                     = "Ivory Coast",
    "Lao People's Democratic Republic"                  = "Laos",
    "Democratic People's Republic of Korea"             = "North Korea",
    "Palestinian Authority or West Bank and Gaza Strip" = "Palestine/West Bank",
    "Congo"                                             = "Republic of the Congo",
    "Syrian Arab Republic"                              = "Syria",
    "Gambia"                                            = "The Gambia",
    "Viet Nam"                                          = "Vietnam"
  )) %>%
  group_by(Recipient, TIME_PERIOD) %>%
  summarise(total_oda_commit_usd_millions = sum(OBS_VALUE, na.rm = TRUE), .groups = "drop") %>%
  rename(country_name = Recipient, year = TIME_PERIOD)
```

### Merge with V-Dem and World Bank data

```{r}
analysis_commit <- vdem %>%
  left_join(oecd_commit, by = c("country_name", "year")) %>%
  arrange(country_name, year) %>%
  group_by(country_name) %>%
  mutate(delta_polyarchy = v2x_polyarchy - lag(v2x_polyarchy)) %>%
  ungroup() %>%
  filter(year >= 1966, year <= 2024) %>%
  dplyr::select(
    country_name, year, total_oda_commit_usd_millions,
    v2x_polyarchy, delta_polyarchy,
    v2x_corr, v2x_rule, v2xcl_rol
  ) %>%
  left_join(wdi, by = c("country_name", "year")) %>%
  mutate(oda_commit_count = pmax(0, round(total_oda_commit_usd_millions))) %>%
  dplyr::select(-total_oda_commit_usd_millions) %>%
  filter(!is.na(oda_commit_count), !is.na(v2x_polyarchy))

write_csv(analysis_commit, here("sania/analysis_commitments.csv"))
```

### Overdispersion check

```{r}
tibble(
  mean     = mean(analysis_commit$oda_commit_count, na.rm = TRUE),
  variance = var(analysis_commit$oda_commit_count,  na.rm = TRUE),
  ratio    = var(analysis_commit$oda_commit_count,  na.rm = TRUE) / mean(analysis_commit$oda_commit_count, na.rm = TRUE)
)
```

### Backsliding model: Commitments (Negative Binomial)

```{r}
nb_commit <- glm.nb(
  oda_commit_count ~ delta_polyarchy + v2x_corr + v2x_rule + v2xcl_rol +
                     log(gdp_per_capita) + log(population),
  data = analysis_commit
)

summary(nb_commit)
```

**Results:** n = 6,039 country-years (868 dropped due to missingness). Theta = 1.174. Residual deviance = 6,852 on 6,031 df. AIC = 87,640.

| Predictor | Estimate | p-value | Interpretation |
|------------------|------------------|------------------|------------------|
| `delta_polyarchy` | −0.285 | .276 | **Not significant.** Backsliding weakly associated with less committed aid, but indistinguishable from zero |
| `v2x_polyarchy` | −0.260 | .007 | More democratic countries receive fewer aid commitments |
| `v2x_corr` | 1.260 | \<.001 | Higher corruption associated with more commitments |
| `v2x_rule` | 1.623 | \<.001 | Better rule of law associated with more commitments |
| `v2xcl_rol` | 0.622 | \<.001 | Stronger civil liberties associated with more commitments |
| `log(gdp_per_capita)` | −0.267 | \<.001 | Richer countries receive fewer commitments |
| `log(population)` | 0.560 | \<.001 | Larger countries receive more commitments |

**Interpretation:** The null result on backsliding holds using commitment data (est. = −0.285, p = .276). The pattern across controls is nearly identical to the disbursement model, suggesting the finding is not an artifact of the outcome measure. Donors do not appear to respond to year-to-year democratic change whether measured through pledged or transferred aid.

### Robustness: Alternative democracy measures (Commitments)

```{r}
analysis_commit_robust <- merged %>%
  arrange(country_name, year) %>%
  group_by(country_name) %>%
  mutate(
    delta_libdem    = v2x_libdem - lag(v2x_libdem),
    delta_partipdem = v2x_partipdem - lag(v2x_partipdem)
  ) %>%
  ungroup() %>%
  filter(year >= 1966, year <= 2024) %>%
  dplyr::select(country_name, year, v2x_libdem, v2x_partipdem,
                delta_libdem, delta_partipdem) %>%
  left_join(analysis_commit, by = c("country_name", "year"))

nb_commit_libdem <- glm.nb(
  oda_commit_count ~ delta_libdem + v2x_corr + v2x_rule + v2xcl_rol +
                     log(gdp_per_capita) + log(population),
  data = analysis_commit_robust
)

nb_commit_partipdem <- glm.nb(
  oda_commit_count ~ delta_partipdem + v2x_corr + v2x_rule + v2xcl_rol +
                     log(gdp_per_capita) + log(population),
  data = analysis_commit_robust
)

summary(nb_commit_libdem)
summary(nb_commit_partipdem)
```

**Results:**

| Model                                       | Delta estimate | p-value |
|---------------------------------------------|----------------|---------|
| Electoral democracy (`delta_polyarchy`)     | −0.285         | .276    |
| Liberal democracy (`delta_libdem`)          | −0.416         | .207    |
| Participatory democracy (`delta_partipdem`) | −0.348         | .448    |

The null result on backsliding is robust across all three democracy indices in the commitment data. All delta estimates are negative --- consistent with the expectation that backsliding leads to less aid --- but none are statistically distinguishable from zero. This mirrors the pattern in the disbursement analysis and reinforces the conclusion that donors do not respond to year-to-year democratic change regardless of how democracy or aid is measured.

## Fixed effects

Load data

```{r}
analysis = read_csv("analysis_merged.csv")
analysis_commit = read_csv("analysis_commitments.csv")

```

Disbursements outcome with fixed effects

removed v2x_polyarchy

```{r}
library(fixest)
fe_nb = fenegbin(
  oda_count ~ delta_polyarchy + v2x_corr + v2x_rule + v2xcl_rol + 
    log(gdp_per_capita) + log(population) | country_name,
  data = analysis
)

summary(fe_nb)
```

Commitments outcome with fixed effects

```{r}
nb_commit <- fenegbin(
  oda_commit_count ~ delta_polyarchy + v2x_corr + v2x_rule + v2xcl_rol +
                     log(gdp_per_capita) + log(population),
  data = analysis_commit
)
```

