---
title: "Democratic Backsliding and Foreign Aid"
author: "Sania Bala"
date: today
format: html
---

## Setup

Loading required packages. Any missing packages are installed automatically.

```{r}
#| message: false
#| warning: false

required_packages <- c("tidyverse", "WDI", "countrycode", "MASS", "here")
missing_packages <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if (length(missing_packages) > 0) install.packages(missing_packages, repos = "https://cloud.r-project.org")

library(tidyverse)
library(WDI)
library(countrycode)
library(here)
# MASS is loaded later, just before the models, to avoid it masking dplyr::select
```

## Data

### Load raw data

We use two datasets:

- **V-Dem** (Varieties of Democracy): country-year measures of democracy, governance, and civil liberties
- **OECD DAC2A**: total ODA (Official Development Assistance) disbursements by recipient country and year, 1960–2024

```{r}
#| message: false

vdem <- read_csv(here("VDEM.csv"))
oecd_raw <- read_csv(here("sania/full_oecd_D.csv"))
```

### Clean and aggregate OECD data

Country names differ between the two datasets, so we standardize OECD recipient names to match V-Dem before merging. We then aggregate from donor-level rows to total ODA received per recipient country per year.

```{r}
oecd_clean <- oecd_raw %>%
  mutate(Recipient = recode(Recipient,
    "Myanmar"                                           = "Burma/Myanmar",
    "Cabo Verde"                                        = "Cape Verde",
    "China (People's Republic of)"                      = "China",
    "Côte d'Ivoire"                                     = "Ivory Coast",
    "Lao People's Democratic Republic"                  = "Laos",
    "Democratic People's Republic of Korea"             = "North Korea",
    "Palestinian Authority or West Bank and Gaza Strip" = "Palestine/West Bank",
    "Congo"                                             = "Republic of the Congo",
    "Syrian Arab Republic"                              = "Syria",
    "Gambia"                                            = "The Gambia",
    "Viet Nam"                                          = "Vietnam"
  ))

oecd <- oecd_clean %>%
  group_by(Recipient, TIME_PERIOD) %>%
  summarise(total_oda_usd_millions = sum(OBS_VALUE, na.rm = TRUE), .groups = "drop") %>%
  rename(country_name = Recipient, year = TIME_PERIOD)

write_csv(oecd, here("sania/oecd_clean.csv"))
```

### Merge V-Dem and OECD

We merge on `country_name` and `year`, keeping all V-Dem observations (left join). We also compute `delta_polyarchy` — the year-over-year change in the electoral democracy index — **before** subsetting years, so that the first observation in the analysis window can draw its lag from the prior year.

- Negative `delta_polyarchy` = democratic backsliding
- Positive `delta_polyarchy` = democratization

```{r}
merged <- vdem %>%
  left_join(oecd, by = c("country_name", "year")) %>%
  arrange(country_name, year) %>%
  group_by(country_name) %>%
  mutate(delta_polyarchy = v2x_polyarchy - lag(v2x_polyarchy)) %>%
  ungroup()
```

### Subset to analysis variables (1960–2024)

We keep only the variables needed for analysis and filter to the period covered by the OECD data.

| Variable | Description |
|---|---|
| `v2x_polyarchy` | Electoral democracy level (control) |
| `delta_polyarchy` | Year-over-year change in democracy (key predictor) |
| `v2x_corr` | Political corruption index |
| `v2x_rule` | Rule of law index |
| `v2xcl_rol` | Civil liberties / equality before the law |

```{r}
analysis <- merged %>%
  filter(year >= 1960, year <= 2024) %>%
  dplyr::select(
    country_name, year, total_oda_usd_millions,
    v2x_polyarchy, delta_polyarchy,
    v2x_corr, v2x_rule, v2xcl_rol
  )
```

### Add GDP per capita and population

GDP per capita and population are not in our V-Dem extract, so we pull them from the World Bank via the `WDI` package.

```{r}
#| message: false
#| warning: false

wdi <- WDI(
  indicator = c(gdp_per_capita = "NY.GDP.PCAP.KD",
                population     = "SP.POP.TOTL"),
  start = 1960, end = 2024,
  extra = FALSE
) %>%
  mutate(country_name = countrycode(iso2c, origin = "iso2c",
                                    destination = "country.name")) %>%
  filter(!is.na(country_name)) %>%
  dplyr::select(country_name, year, gdp_per_capita, population)

analysis <- analysis %>%
  left_join(wdi, by = c("country_name", "year"))
```

### Prepare outcome variable

ODA is reported in USD millions (continuous). We convert it to an integer count by rounding to the nearest million and flooring any negative values (loan repayments) to zero, as required by count models.

```{r}
analysis <- analysis %>%
  mutate(oda_count = pmax(0, round(total_oda_usd_millions))) %>%
  dplyr::select(-total_oda_usd_millions) %>%
  filter(!is.na(oda_count), !is.na(v2x_polyarchy))


write_csv(analysis, here("sania/analysis_merged.csv"))
```

## Models

### Model strategy

We use negative binomial regression to test whether democratic backsliding predicts foreign aid received. ODA is heavily overdispersed — its variance far exceeds its mean — which rules out Poisson. Population is included as a control (`log(population)`).

```{r}
#| message: false
library(MASS)
```

### Overdispersion check

Under Poisson, the variance should equal the mean. If the variance-to-mean ratio is much greater than 1, the data is overdispersed and negative binomial is more appropriate.

```{r}
tibble(
  mean     = mean(analysis$oda_count, na.rm = TRUE),
  variance = var(analysis$oda_count,  na.rm = TRUE),
  ratio    = var(analysis$oda_count,  na.rm = TRUE) / mean(analysis$oda_count, na.rm = TRUE)
)
```

The variance-to-mean ratio is far greater than 1, confirming severe overdispersion. We use negative binomial regression, which adds a dispersion parameter to account for this.

### Backsliding model: Democratic backsliding and aid (Negative Binomial)

The key predictor is `delta_polyarchy` (year-over-year change in `v2x_polyarchy`). We keep `v2x_polyarchy` as a level control so that `delta_polyarchy` captures the effect of change net of overall democratic standing. The expected sign on `delta_polyarchy` is positive — backsliding (negative delta) should predict less ODA.

```{r}
nb_backsliding <- glm.nb(
  oda_count ~ delta_polyarchy + v2x_polyarchy + v2x_corr + v2x_rule + v2xcl_rol +
              log(gdp_per_capita) + log(population),
  data = analysis
)

summary(nb_backsliding)
```

**Results:** *(figures below reflect the 2010–2020 window; re-run to update for the full 1960–2024 series)* n = 1,216 country-years. AIC = 26,429 (marginally better than Model 1).

| Predictor | Estimate | p-value | Interpretation |
|---|---|---|---|
| `delta_polyarchy` | 0.466 | .537 | **Not significant.** Direction is correct but effect is indistinguishable from zero |
| `v2x_polyarchy` | −1.001 | \<.001 | Democracy level remains strongly significant |
| All controls | — | — | Consistent with Model 1 |

**Interpretation:** The backsliding hypothesis is not supported in this specification. Aid allocations respond to a country's overall democratic standing, but not to year-to-year democratic change. This likely reflects donor inertia — aid budgets are planned in advance and respond slowly to political shifts. Adding fixed effects or a longer lag structure (2–3 years) may surface a delayed donor response.

## Robustness Checks

### Check 1: Alternative democracy measures

`v2x_polyarchy` measures electoral democracy specifically. If the null result is real, it should hold across different conceptualizations of democracy. We test two alternatives:

- `v2x_libdem` — liberal democracy (adds judicial constraints and civil liberties)
- `v2x_partipdem` — participatory democracy (emphasizes civil society and direct participation)

```{r}
analysis_robust <- merged %>%
  arrange(country_name, year) %>%
  group_by(country_name) %>%
  mutate(
    delta_libdem    = v2x_libdem - lag(v2x_libdem),
    delta_partipdem = v2x_partipdem - lag(v2x_partipdem)
  ) %>%
  ungroup() %>%
  filter(year >= 1960, year <= 2024) %>%
  dplyr::select(country_name, year, v2x_libdem, v2x_partipdem,
                delta_libdem, delta_partipdem) %>%
  left_join(analysis, by = c("country_name", "year"))

nb_libdem <- glm.nb(
  oda_count ~ delta_libdem + v2x_libdem + v2x_corr + v2x_rule + v2xcl_rol +
              log(gdp_per_capita) + log(population),
  data = analysis_robust
)

nb_partipdem <- glm.nb(
  oda_count ~ delta_partipdem + v2x_partipdem + v2x_corr + v2x_rule + v2xcl_rol +
              log(gdp_per_capita) + log(population),
  data = analysis_robust
)

summary(nb_libdem)
summary(nb_partipdem)
```

**Results:** The null result holds across both alternatives.

| Model | Delta estimate | p-value |
|---|---|---|
| Electoral democracy (`delta_polyarchy`) | 0.466 | .537 |
| Liberal democracy (`delta_libdem`) | 0.185 | .840 |
| Participatory democracy (`delta_partipdem`) | 1.326 | .266 |

The null result on backsliding is robust across all three conceptualizations of democracy. This consistently points to donor inertia — aid allocations respond to where a country sits on the democracy scale, not to how fast it is moving.

### Check 2: Extended time window

The analysis now uses the full OECD time series (1960–2024), providing a much longer panel with more backsliding episodes and greater statistical power compared to the original 2010–2020 window. Year coverage in practice is limited by the availability of V-Dem scores and World Bank GDP/population data, but the full window is retained and listwise deletion handles missing observations.

## Potential Improvements

1. **Lag the predictors** — currently democracy scores and ODA are measured in the same year, raising reverse causality concerns. Lagging `v2x_polyarchy` and all controls by one year (t-1) would make the causal direction cleaner.

2. **Country + year fixed effects** — country fixed effects absorb all time-invariant country characteristics (colonial history, geographic proximity to donors). Year fixed effects absorb global shocks. Can be implemented with the `fixest` package.

3. **Check for zero inflation** — if many country-years have zero ODA, a zero-inflated negative binomial (ZINB) may fit better. Check with `table(analysis$oda_count == 0)`.
