---
title: "Democratic Backsliding and Foreign Aid"
author: "Joe Hoskisson and Sania Bala"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    toc: false
    smaller: true
    css: styles.css
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

# load packages and data — runs silently before any slide
library(tidyverse)
library(MASS)
library(knitr)
library(here)

# load pre-built analysis dataset (produced by data_merge.qmd)
# here() always resolves to the project root (.Rproj location)
analysis <- read_csv(here("sania/analysis_merged.csv"))

nb_backsliding <- glm.nb(
  oda_count ~ delta_polyarchy + v2x_polyarchy + v2x_corr + v2x_rule + v2xcl_rol +
              log(gdp_per_capita) + log(population),
  data = analysis
)
```

## Research Question & Motivation

**What are we trying to learn?**

> Does democratic backsliding cause countries to receive less foreign aid?

**Why does it matter?**

- Democratization. Liberalism view would require certain reforms to get aid.
- Punishment. Do donors hold governments accountable when they violate democratic norms?
- Reward. Does more OECD aid go to governments that are becoming more democratic?

---

## Data

**Sources**

| Dataset | What it provides |
|---|---|
| V-Dem (Coppedge et al.) | Country-year democracy indices and governance measures |
| OECD DAC2A | Total ODA disbursements by recipient country and year |
| World Bank WDI | GDP per capita and population |

*ODA (Official Development Assistance) refers to total aid received by developing countries from donor governments. It includes grants, concessional loans, and technical assistance.*

**Sample**

- Unit of analysis: **country-year**
- Period: **2010–2020**
- Final N: **`r nrow(analysis)` country-years** 

---

## Key Variables

```{r}
#| label: variable-table

var_table <- tribble(
  ~Role,          ~Variable,            ~Description,
  "Outcome",      "oda_count",          "Total ODA received (USD millions, rounded)",
  "Key predictor","delta_polyarchy",    "Year-over-year change in electoral democracy (V-Dem)",
  "Control",      "v2x_polyarchy",      "Electoral democracy level (V-Dem)",
  "Control",      "v2x_corr",           "Political corruption index (V-Dem)",
  "Control",      "v2x_rule",           "Rule of law index (V-Dem)",
  "Control",      "v2xcl_rol",          "Civil liberties / equality before the law (V-Dem)",
  "Control",      "log(gdp_per_capita)","Log GDP per capita, constant 2015 USD (World Bank)",
  "Control",      "log(population)",    "Log population (World Bank)"
)

kable(var_table, align = "lll")
```


---

## Outcome & Model Choice

**What kind of outcome do we have?**

- ODA is a **non-negative count** (USD millions, rounded to nearest integer)
- Highly right-skewed — a small number of countries receive enormous amounts; most receive relatively little
- **Overdispersion:** variance far exceeds the mean, which violates the Poisson assumption

**Which model are we fitting?**

- **Negative binomial regression** — extends Poisson by adding a dispersion parameter (θ) that absorbs excess variance
- Confirmed: Poisson residual deviance = 36,501,834 on 1,211 df; NB residual deviance = 1,425 on 1,211 df

---

## Preliminary Results: Coefficient Table


```{r}
#| label: coef-table

# extract and format coefficients from both models
fmt <- function(model) {
  tidy <- broom::tidy(model)
  tidy %>%
    filter(term != "(Intercept)") %>%
    transmute(
      Predictor = recode(term,
        "delta_polyarchy"    = "Backsliding (Δ democracy)",
        "v2x_polyarchy"      = "Democracy level",
        "v2x_corr"           = "Corruption",
        "v2x_rule"           = "Rule of law",
        "v2xcl_rol"          = "Civil liberties",
        "log(gdp_per_capita)"= "log(GDP per capita)",
        "log(population)"    = "log(Population)"
      ),
      Estimate  = round(estimate, 3),
      `Std. Error` = round(std.error, 3),
      `p-value` = case_when(
        p.value < .001 ~ "<.001",
        TRUE           ~ as.character(round(p.value, 3))
      )
    )
}

kable(fmt(nb_backsliding), align = "lrrr")
```

---

## Preliminary Results: Predicted Values Plot

```{r}
#| label: pred-plot
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.5

# build a prediction grid: vary delta_polyarchy, hold everything else at median
pred_data <- tibble(
  delta_polyarchy  = seq(min(analysis$delta_polyarchy, na.rm = TRUE),
                         max(analysis$delta_polyarchy, na.rm = TRUE),
                         length.out = 100),
  v2x_polyarchy    = median(analysis$v2x_polyarchy,    na.rm = TRUE),
  v2x_corr         = median(analysis$v2x_corr,         na.rm = TRUE),
  v2x_rule         = median(analysis$v2x_rule,         na.rm = TRUE),
  v2xcl_rol        = median(analysis$v2xcl_rol,        na.rm = TRUE),
  gdp_per_capita   = median(analysis$gdp_per_capita,   na.rm = TRUE),
  population       = median(analysis$population,       na.rm = TRUE)
)

preds <- predict(nb_backsliding, newdata = pred_data, type = "link", se.fit = TRUE)
pred_data <- pred_data %>%
  mutate(
    predicted_oda = exp(preds$fit),
    lower         = exp(preds$fit - 1.96 * preds$se.fit),
    upper         = exp(preds$fit + 1.96 * preds$se.fit)
  )

ggplot(pred_data, aes(x = delta_polyarchy, y = predicted_oda)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "maroon", alpha = 0.2) +
  geom_line(linewidth = 1.2, color = "maroon") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey30") +
  annotate("text", x = -0.01, y = max(pred_data$predicted_oda) * 0.95,
           label = "← Backsliding", hjust = 1, color = "grey30", size = 3.5) +
  annotate("text", x =  0.01, y = max(pred_data$predicted_oda) * 0.95,
           label = "Democratization →", hjust = 0, color = "grey30", size = 3.5) +
  labs(
    title    = "Predicted ODA by Democratic Change",
    subtitle = "All other variables held at median",
    x        = "Δ Electoral Democracy (delta_polyarchy)",
    y        = "Predicted ODA (USD millions)"
  ) +
  theme_minimal(base_size = 13)
```

---

## Interpreting the Results

- `delta_polyarchy` is **not significant** (est. = 0.299, p = .618)
- Direction is correct — backsliding predicts less aid, democratization predicts more
- But the effect is too small and uncertain to distinguish from zero
- **Substantive interpretation:** donors respond to *where* a country sits on the democracy scale, not to *how fast* it is moving 

---

## What We're Stuck On & What's Next

**Things we are still working through**

- Should we test **alternative democracy indices** (e.g., liberal, participatory) in the VDEM dataset to assess whether results are sensitive to how democracy is measured?

- Should we **extend the backsliding measure** to capture cumulative democratic decline over a 3–5 year window, rather than year-over-year change, to better reflect how donors perceive and respond to backsliding?

- Should we look at the differences between **commitments** and **disbursement** of aid as democracy ratings change?
